Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance in zooplankton

### 2022-06-12 text/Manuscript.Rmd
MS: results 1 edits

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data. Phylogenetic analysis of NOS proteins indicate that the *Platynereis* NOS belongs to an orthology group of metazoan NOS sequences [need to add tree]. To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis. In two-day-old larvae, we detected *NOS* expression in four cells in the apical organ region (Fig. 1A). The same four cells cells were labelled with a reporter construct driving palmitoylated-tdTomato by a 12 kbp upstream genomic region of the *NOS* gene. This reporter also revealed the axonal projections of the NOS-expressing neurons. The position and axonal projections of the four *NOS*-expressing cells allowed us to identify the same cells as four interneurons in the ciliary-type photoreceptor cells (cPRCs) circuit from Platynereis reconstruction data (Fig. 1B). We named these NOS interneurons, INNOS (INNOSdl, dr, vl, vr) (Fig. 1C). The cPRC circuit has been reported as a neural circuit involved in c-opsin1 mediated UV avoidance behaviour (Verasztó et al., 2017). INNOS were presynaptic from the cPRCs and postsynaptic to the RGW interneurons in the cPRC circuit (Fig. 1D). To determine whether NO is involved in UV avoidance behaviour, we generated mutant lines with the CRISPR/Cas-9 system and examined their behavioural responses to UV light.

>>>>

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data. Phylogenetic analysis of NOS proteins indicate that the *Platynereis* NOS belongs to an orthology group of metazoan NOS sequences [@Kei - need to add tree]. To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis. In two-day-old larvae, we detected *NOS* expression in four cells in the apical organ region (Fig. 1A) [@Kei - 3d HCR and transgene?]. The same four cells cells were labelled with a reporter construct driving palmitoylated-tdTomato by a 12 kbp upstream genomic region of the *NOS* gene. This reporter also revealed the axonal projections of the NOS-expressing neurons. The position and axonal projections of the four *NOS*-expressing cells allowed us to identify the same cells in the three-day-old connectome data as four interneurons (INNOS) in the ciliary photoreceptor-cell (cPRC) circuit (Figure 1D-E). The cPRCs express the UV-absorbing c-opsin1 and mediate a UV avoidance behaviour in the larvae [@Verasztó2018] and modulates locomotion and maturation in adult worms [@Veedin2021]. INNOS cells are postsynaptic to the cPRCs and presynaptic to the RGW interneurons in the cPRC circuit (Fig. 1F).

To determine whether NO is involved in UV avoidance behaviour, we generated mutant lines with the CRISPR/Cas-9 system and examined their behavioural responses to UV light.
